### 2020-12-11 paper/2020-10-revision-background.tex
Shorten version of Conceptual Framework and Introduction

This section provides a conceptual framework for understanding how birth spacing changes
and guiding the subsequent analysis.
I first introduce three potential explanations that link fertility and birth 
spacing decisions: economic conditions, desired child outcomes, and son preference 
\citep{Casterline2016,Portner2018}.
Next, I discuss why female education and area of residence are the principle factors 
in the empirical analyses and how they tie into the three explanations.

% Economic conditions
The improvements in economic condition, especially the doubling of real wages,
are likely to affect fertility although the direction is ambiguous, as, empirically, 
higher female wages lowers fertility, whereas higher male wages increases fertility 
\citep{Hotz1997,schultz97}.
Economic theory predicts that the effect of higher wages on fertility is determined by 
the relative strengths of the substitution and income effects.
The substitute effect captures that as wages increase time becomes more expensive, and
people therefore work more and spend less time on non-wage earning activities, such as 
children or leisure.
The income effect captures that higher wages increase the available income, which
leads people to spend more time on time-intensive activities and less time working.
Since women spend substantially more time on childrearing than men, the substitution effect 
dominates for women but the income effect for men.

If having children requires the mother to curtail her economic activities, higher
female wages, in addition to lowering fertility, will tend to shorten birth spacing.
Shortening birth spacing allows parents to take advantage of economies of scale in 
childrearing---for example, looking after two children simultaneously requires 
less than double the time needed to look after one child \citep{Vijverberg1982,Hotz1997}.

% Child outcomes
Parents care both about how many children they have and their outcomes, and with 
increasing returns to education and lower offspring mortality, parents are likely to 
reduce fertility and invest more in each child \citep{Rosenzweig1982a,Wolpin1997}.
The higher return to education means that parents have more of an incentive to 
invest in their children's education, which increases the cost of children and 
lowers fertility. 
Lower expected mortality means that fewer births are required to reach a 
desired number of surviving children and allow parents to invest more in each child.

Longer spacing between births is generally expected to have positive effects on 
child outcomes and if fertility falls because of increased desired for better child
outcomes we would also expect spacing to increase.
The clearest example of birth spacing affecting child outcomes is for health,
where very short spacing---approximately 24 months or less---leads to worse health 
and mortality outcomes for children 
\citep{Whitworth2002,Conde-Agudelo2012}.
Whether longer spacing also results in better human capital outcome is unclear as the 
evidence is mixed for developed countries and nonexistent for developing countries
\citep{Zajonc1976,Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

% Son preference 
Empirically, a stronger son preference has only a minor effect on average fertility, but
son preference still affects the sex composition and fertility across families through 
differential stopping behavior and leads to shorter spacing and  worse health outcomes for 
daughters when there are no sons in the household 
\citep{repetto72,clark00,Whitworth2002,Basu2010,Jayachandran2011,Barcellos2014}.

Another result of son preference is the use of sex selection, which enables parents to 
avoid giving birth to a girl but increases the expected interval to the next birth
by 6--12 months for each abortion.
This increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover, 
or the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
Second, the waiting time to conception is one to six months. 
Finally, sex determination tests are reliable only from three months of gestation.

The use of sex selection, and therefore the length of birth intervals, increases with
lower desired fertility and the higher the parity provided the desired number of sons has 
not been reached \citep{Portner2015b,Jayachandran2017}.

% Factors/predictions
Although the explanations above are based on economic conditions, desired child outcome,
son preference, and the access to sex selection, none of these are available in the
data, so I instead use female education and area of residence as the two main explanatory 
variables in the empirical analyses.
Female education and area of residence are the closest available proxies and earlier 
research show that both higher female education and urbanization are associated with 
lower fertility and increased use of sex selection, and, therefore, likely also changes 
in birth spacing
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Furthermore, both variables are interesting given the tremendous increases in 
both education and urbanization in India; 
Figure \ref{fig:education_over_time} show the substantial increase in female education
in both rural and urban areas, although it is still considerably higher in urban than in 
rural areas and the proportion of the population living in urban areas almost doubled from 
18\% in 1960 to 35\% in 2019 \citep{United-Nations2019}.

@@

Women living in urban areas are likely to have lower fertility and higher use
of sex selection than women in rural areas.
The lower fertility is the result of higher cost of children and lower child mortality 
risk in urban compared to rural areas.%
\footnote{
However, holding parental education constant, there is evidence that children in
Indian slums do worse than they would have in rural areas \citep{Portner2018a}.
}
The higher use of sex selection comes partly from the lower fertility and partly from 
women in urban areas having higher income and better access to health care facilities.

As women receive more education their potential wages increase, which should shorten 
birth spacing and lower fertility, but the low and declining female labor force 
participation in India, especially for younger women, suggests that families face little 
incentive to space children more closely together for economic reasons.
Even as women's education increased, their labor force participation stagnated or 
decreased and is now one of the lowest outside the Middle East 
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018}.%
\footnote{
In line with other countries, there is still a U-shaped relationship between education and 
labor force participation for married women, with the lowest participation for 
women with 8--11 years of education \citep{Goldin1994,Chatterjee2018}.
See appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}

One explanation for the low female labor force participation is that male wages have 
increased so substantially that the income effect dominates any substitution effect.
The average male wage is almost 70\% higher than the female wage and women’s labor 
supply appears to be more negatively elastic to husbands' wages than positively elastic 
to their own wages \citep{Klasen2015,Bhargava2018}.

Furthermore, the increasing return to male education may make it appear as if higher 
female education results in lower fertility and more sex selection, if women with more 
education are better at producing child human capital as suggested by \citet{Behrman1999},.
The higher return to male education provides an additional incentive to invest in sons'
education, and one such investment is a better-educated mother.
As discussed above, the higher returns should also lower desired fertility, making the use 
of sex selection even more appealing, and possibly increases birth spacing.

% Sanskritization
Although the expansion in female education might bring in groups with initially different 
norms, the process of ``Sanskritization'' implies that as lower-castes females gain access 
to education, they adopt higher-caste norms such as stronger son preference and a retraction 
from the formal labor market \citep{Srinivas1956}.
The low and declining female labor force participation as husbands' wages increase 
indicates that this process still operates \citep{Abraham2013,Chatterjee2018}.

% Summary / hypotheses

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect longer birth spacing over time, independent of education 
levels.
Furthermore, birth spacing likely increases the most among the better educated 
because their household income increases the most---even with declining female labor 
participation---and because of their use of sex selection.
Finally, even with the substantial increase in the number of better-educated women, 
"Sanskritization" implies that the changing composition will not substantially change the 
use of sex selection.

>>>>

This section provides a conceptual framework for understanding changes in birth spacing.
I first introduce three potential explanations that link fertility and birth 
spacing decisions: economic conditions, desired child outcomes, and son preference 
\citep{Casterline2016,Portner2018}.
Next, I discuss why female education and area of residence are the principal factors 
in the empirical analyses and how they tie into the three explanations.

% Economic conditions
The improvements in economic condition, especially the doubling of real wages,
are likely to affect fertility although the direction is ambiguous; empirically, 
higher female wages lowers fertility, whereas higher male wages increases fertility 
\citep{Hotz1997,schultz97}.
Economic theory predicts that the effect of higher wages on fertility is determined by 
the relative strengths of the substitution and income effects.
The substitute effect captures that as wages increase time becomes more expensive, and
people therefore work more and spend less time on non-wage earning activities, such as 
children or leisure.
The income effect captures that higher wages increase the available income, which
leads people to spend more time on time-intensive activities and less time working.
Since women spend substantially more time on childrearing than men, the substitution effect 
dominates for women but the income effect for men.

Higher female wages will also tend to shorten birth spacing if having children requires 
the mother to curtail her economic activities.
Shortening birth spacing allows parents to take advantage of economies of scale in 
childrearing---for example, looking after two children requires less than double the time 
needed for one child \citep{Vijverberg1982,Hotz1997}.

% Child outcomes
Parents care both about how many children they have and their outcomes, and with 
increasing returns to education and lower offspring mortality, parents are likely to 
reduce fertility and invest more in each child \citep{Rosenzweig1982a,Wolpin1997}.
The higher return to education means a stronger incentive to invest in their children's
education, which increases the cost of children and lowers fertility. 
With lower expected mortality fewer births are required to reach a desired number 
of surviving children, which furthermore allows parents to invest more in each child.

An increased desired for better child outcomes will also tend to lengthen birth spacing,
provided it has positive effects on child outcomes.
The clearest example of a positive effect of longer spacing that very short 
spacing---approximately 24 months or less---leads to worse child health and mortality 
outcomes \citep{Whitworth2002,Conde-Agudelo2012}.
Whether longer spacing also improves human capital outcomes is more speculative with
mixed evidence for developed countries and no evidence for developing countries
\citep{Zajonc1976,Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

% Son preference 
Empirically, stronger son preference only marginally increases average fertility, but
differential stopping behavior still affects the sex composition and fertility across 
families and leads to shorter spacing and worse health outcomes for daughters  
when there are no sons in the household
\citep{repetto72,clark00,Whitworth2002,Basu2010,Jayachandran2011,Barcellos2014}.

Stronger son preference increases the use of sex selection and each sex-selective
abortion increases the expected birth spacing by 6--12 months.
After an abortion the uterus needs at least two menstrual cycles---approximately 
two months---to recover, or the subsequent risk of spontaneous abortion increases 
substantially \citep{zhou00b}.
Once conception can be attempted, the waiting time to conception is likely between
one and six months. 
Finally, sex-determination tests are reliable only from three months of gestation. 
Hence, it takes at least six months before the couple is at the same point in the next 
pregnancy as they were in the prior pregnancy when they decided to abort.

The use of sex selection, and therefore the length of birth intervals, increases with
lower desired fertility and with parity, provided the desired number of sons has 
not been reached \citep{Portner2015b,Jayachandran2017}.

% Factors/predictions
Although the explanations above are based on economic conditions, desired child outcomes,
son preference, and sex selection, none of these are available in the data, so I instead 
use female education and area of residence as the two principal explanatory variables in 
the empirical analyses.
Not only are these the closest available proxies, prior research shows that both higher 
female education and urbanization are associated with lower fertility and increased 
sex selection use, and, therefore, likely also changes birth spacing
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Furthermore, both variables are of interest given the tremendous changes in India; 
Figure \ref{fig:education_over_time} shows the substantial increase in female education
and the urban proportion of the population almost doubled from 18\% in 1960 to 35\% in 
2019 \citep{United-Nations2019}.

@@

Urban women are likely to have lower fertility and higher use of sex selection than women 
in rural areas.
The lower fertility is the result of higher cost of children and lower child mortality 
risk in urban compared to rural areas.%
\footnote{
However, holding parental education constant, children in Indian slums have worse
health outcomes than in rural areas \citep{Portner2018a}.
}
The higher use of sex selection comes partly from the lower fertility and partly from
higher income and better access to health care facilities in urban areas.

As women receive more education their potential wages increase, which should shorten 
birth spacing and lower fertility, but the low and declining female labor force 
participation in India suggests that families face little incentive to space children 
more closely together for economic reasons.
Even as women's education increased, their labor force participation stagnated or 
decreased and is now one of the lowest outside the Middle East 
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018}.

The decline in female labor force participation appear to be driven by a combination
of a relatively small expansion in the sectors where women work and the income effect 
from rapidly increasing male wages and education dominating the substitution effect from 
higher female wages \citep{Klasen2015,Bhargava2018}.
The income effect can dominate because not only do men have a higher average level of 
education than women, men are also paid more than women for a given level of education, 
and the returns to education increases with education in India \citep{Agrawal2011}.
Hence, increasing male education will result in an larger increase in income than a 
similar increase in female education.

However, in a situation with increasing returns to male education and declining female labor 
force participation, we might still observe an association between higher female education 
and both lower fertility and increased use of sex selection, provided better-educated women 
are better at producing child human capital as suggested by \citet{Behrman1999}.
In this scenario, better-educated women are desired not for their ability to generate
income but for their ability to produce better-educated sons.
The higher return to male education provides an additional incentive to invest in sons'
education---with a better-educated mother one such investment---and also lowers desired 
fertility, makes sex selection more likely, and possibly increases birth spacing.

% Sanskritization
Finally, although the expansion in female education will likely bring in groups who 
initially have different norms, the process of ``Sanskritization'' implies that as 
lower-castes females gain access to education, they adopt higher-caste norms such as stronger son 
preference and a retraction from the formal labor market \citep{Srinivas1956}.
The declining female labor force participation suggests that this process still 
operates \citep{Abraham2013,Chatterjee2018}.

% Summary / hypotheses
In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect longer birth spacing over time, independent of education 
levels.
Furthermore, birth spacing likely increases the most among the better educated 
because their household income increases the most---even with declining female labor 
participation---and because of their use of sex selection.
Finally, "Sanskritization" implies that the changing composition of better-educated
women will not substantially change the use of sex selection.

### 2020-12-11 paper/sex_selection_spacing-ver5.tex
Shorten version of Conceptual Framework and Introduction

% Background
India has experienced many positive changes over the past four decades:
The economy has grown substantially,
educational attainment has increased for males and females,
and the total fertility rate has fallen to 2.2 children
\citep{Bosworth2008,Dharmalingam2014,
International-Institute-for-Population-Sciences-IIPS2017}.

@@

% Questions addressed and why valuable
The main question I address here is how birth spacing responded to these changes, 
especially the spread of sex selection. 
The motivation is twofold.
First, researchers have failed to appreciate that each abortion increases the
interval between births by 6--12 months.%
\footnote{
The increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover, 
or the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
Second, the waiting time to conception is one to six months. 
Finally, sex determination tests are reliable only from three months of gestation. 
}
Hence, the growing use of sex-selective abortions may substantially increase birth spacing, 
although we do not know to what extent. 
Second, greater educational attainment by women, higher household income, and the low and 
declining female labor force participation all likely influence birth spacing. 
Thus, the combined changes in birth spacing may outpace what we have observed in 
other countries.

@@

% Combine descriptive and theory sections into one

\section{Background and Conceptual Framework}

To set the stage for the subsequent analyses and provide a conceptual framework for 
understanding birth spacing, I first discuss female education and labor force 
participation in India and relevant theories on birth spacing.

% [Female education]

Female education is a crucial explanatory variable here for three reasons.
First, higher female education is associated with lower fertility and increased use of 
sex selection
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Second, female labor force participation in India, as in other developing countries, 
first decreases and then increases with education
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
Finally, child mortality decreases with maternal education
\citep{Rosenzweig1982a,Whitworth2002,Maitra2008}.

% [TK where to put son preference discussion?]
% %
% \footnote{
% It is also possible that the effect is driven by a stronger son preference, although 
% higher education women have shown declining son preference \citep{bhat03,pande07}.
% }

I divide education levels into four groups: No education, 1--7 years, 8--11 years, and 
12 and more years.
The latter two correspond to having completed primary and secondary school, 
respectively.%
\footnote{
Although there are variations by state, elementary education in India consists of a 
primary school covering grades one through five and an upper primary---or middle 
school---covering grades six through eight. 
Similarly, secondary education covers grades nine and tenth for ``secondary education'' 
and 11 and 12 for ``upper secondary.''
}
To ensure that the results are comparable with the prior literature on fertility
and mortality in India, I follow the NFHS reports, except that I combine the 
less than five years and 5--7 years of schooling completed and the 8--9 
and 10--11 years of schooling completed.
This grouping allows me to capture the differences across education levels discussed below, 
while also having groups large enough for the empirical method.

% [what has happened to education]
Female education has increased substantially over time in rural and urban areas but 
is still substantially higher in urban than in rural areas.
Figure \ref{fig:education_over_time} shows the distribution of schooling by birth 
cohort for urban and rural women, 
20 years or older, whether married or not, based on the four rounds of the NFHS.
In rural areas, women with no education have gone from 90\% for the 1930s cohorts to 
less than 20\% for the 1990s cohorts.
Women with eight or more years of education have gone from almost zero for the 1930 cohort 
to more than 60\% for the 1990s cohorts, with about half in the 8--11 group and the 
other half in the 12 plus group.
In urban areas, the proportions with no education or 1--7 years have each declined 
to just below 10\%.
Most of the increase in urban female education came from the 12 plus group, which 
now accounts for more than half of all urban women for the most recent cohorts.

\begin{figure}[htpb]
\centering
\subfloat[Rural]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_rural}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_urban} 
    \end{minipage}
}
\caption{Distribution of education by cohort for women 20 years or older at survey}
\label{fig:education_over_time}
\end{figure}

% [female labor force participation]

The standard economic argument for shorter spacing with increasing female education 
is that parents incur time costs when they have children \citep{Hotz1997,schultz97}.
Suppose having children requires the mother to reduce her market work. 
Parents can then lower this cost of children by shortening birth spacing to take advantage 
of economies of scale in childrearing  \citep{Vijverberg1982}.

However, even as the level of female education has increased, the female labor force 
participation in both urban and rural areas has stagnated or decreased
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
India's female labor force participation is now lower than that of most other countries and does 
not yet show any signs of increasing \citep{Klasen2015,Chatterjee2018}.
In line with previous research, the NFHS data show a U-shaped relationship between 
education and working for married women, with the highest percentage working being women with 
either no education or 12 or more years and the lowest being women with 8--11 years of education.%
\footnote{
See appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}

The low and declining female labor force participation, especially for younger women, 
suggests that families face little incentive to space children more closely 
together for economic reasons.
One explanation is that household income has increased so substantially that the income 
effect dominates any substitution effect.
Two findings speak to this effect.
First, although real wages for both men and women have almost doubled between 1987 and 
2011, the mean male wage is still close to 70\% higher than the female wage 
\citep{Klasen2015,Bhargava2018}.
Second, women’s labor supply appears to be more negatively elastic to husbands' 
wages than positively elastic to their own wages \citep{Bhargava2018}.

% [sex selection]

The introduction of sex selection allows parents to avoid giving birth to girls but 
increases the expected interval to the next birth.
Theory suggests that sex selection increases with lower desired fertility and with
higher parity for a given desired number of children \citep{Portner2015b}.
Sex selection is more widespread among better-educated than less-educated and among urban 
than rural women, which is consistent with lower desired fertility increasing the
use of sex selection 
\citep{das_gupta97,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Better-educated and urban women also tend to live in households with higher income, which 
lowers the relative costs of using sex selection and having long birth intervals.

The combination of rising incomes and continued son preference may lead to even longer 
spacing than might be expected from the income effect alone.
As women's education increases, their productivity in the production of offspring human 
capital also increases.
With relatively more boys born because of increased access to sex-selective 
abortions and the increasing income potential for (male) offspring, demand for 
better-educated women can increase, even if they do not participate in the labor market 
\citep{Behrman1999}.

If more and ``better'' parental attention per child results in higher child ``quality,'' 
we should expect longer birth intervals \citep{Zajonc1975,Zajonc1976,Razin1980}.
However, the evidence on spacing's effect on child quality measures such as IQ 
and education is mixed for developed countries and nonexisting for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
The exception is health and mortality, where longer spacing does lead to better outcomes, 
although this relationship weakens with maternal education 
\citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

% If parents realize that longer spacing leads to better child outcomes, we may even observe 
% later conception, especially when, with sex selection, the next child is likely to be a boy.
% Working in the opposite direction of this "delay" effect is that women with more education 
% can, in principle, space their children closer together without substantially increasing 
% child mortality risk. 
% However, given the apparent lack of pressure to return to the labor force, this effect is 
% likely small.

% Sanskritization

The increases in female educational attainment imply that access to education has 
expanded beyond the higher castes. 
One possible effect of the associated change in the composition of better-educated women 
is that this group's behavior would change.
However, ``Sanskritization'' implies that as lower-castes females gain access to 
education and their husbands' income increases, they adopt higher-caste 
norms such as stronger son preference and a retraction from the formal labor 
market \citep{Srinivas1956,Chen1995,Abraham2013,Chatterjee2018}.
The low and declining female labor force participation suggests that this process still operates.

% Summary / hypotheses

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
Furthermore, I expect birth spacing to increase the most among the better educated 
because their household income increases the most---even with declining female labor 
participation---and because of their use of sex selection.
Even with the substantial increase in the number of better-educated women, 
"Sanskritization" implies that the changing composition will not substantially change 
the use of sex selection.

>>>>

% Background
India has experienced many positive changes over the past four decades:
The economy has grown substantially,
real wages have more than doubled,
educational attainment has increased for males and females,
and the total fertility rate has fallen to 2.2 children
\citep{Bosworth2008,Dharmalingam2014,Klasen2015,
International-Institute-for-Population-Sciences-IIPS2017}.

@@

% Questions addressed and why valuable
The main question I address here is how birth spacing responded to these changes, 
especially the spread of sex selection. 
The motivation is twofold.
First, the growing use of sex-selective abortions may substantially increase birth spacing
because, as detailed below, it takes 6 months or more after an abortion to reach the same 
point in the next pregnancy.
However, researchers have failed to appreciate this effect, and we do not know the size of 
the impact.
Second, greater educational attainment by women, higher household income, and the low and 
declining female labor force participation all likely influence birth spacing. 
Thus, the combined changes in birth spacing may outpace what we have observed in 
other countries.

@@

\section{Conceptual Framework}

This section provides a conceptual framework for understanding changes in birth spacing.
I first introduce three potential explanations that link fertility and birth 
spacing decisions: economic conditions, desired child outcomes, and son preference 
\citep{Casterline2016,Portner2018}.
Next, I discuss why female education and area of residence are the principal factors 
in the empirical analyses and how they tie into the three explanations.

% Economic conditions
The improvements in economic condition, especially the doubling of real wages,
are likely to affect fertility although the direction is ambiguous; empirically, 
higher female wages lowers fertility, whereas higher male wages increases fertility 
\citep{Hotz1997,schultz97}.
Economic theory predicts that the effect of higher wages on fertility is determined by 
the relative strengths of the substitution and income effects.
The substitute effect captures that as wages increase time becomes more expensive, and
people therefore work more and spend less time on non-wage earning activities, such as 
children or leisure.
The income effect captures that higher wages increase the available income, which
leads people to spend more time on time-intensive activities and less time working.
Since women spend substantially more time on childrearing than men, the substitution effect 
dominates for women but the income effect for men.

Higher female wages will also tend to shorten birth spacing if having children requires 
the mother to curtail her economic activities.
Shortening birth spacing allows parents to take advantage of economies of scale in 
childrearing---for example, looking after two children requires less than double the time 
needed for one child \citep{Vijverberg1982,Hotz1997}.

% Child outcomes
Parents care both about how many children they have and their outcomes, and with 
increasing returns to education and lower offspring mortality, parents are likely to 
reduce fertility and invest more in each child \citep{Rosenzweig1982a,Wolpin1997}.
The higher return to education means a stronger incentive to invest in their children's
education, which increases the cost of children and lowers fertility. 
With lower expected mortality fewer births are required to reach a desired number 
of surviving children, which furthermore allows parents to invest more in each child.

An increased desired for better child outcomes will also tend to lengthen birth spacing,
provided it has positive effects on child outcomes.
The clearest example of a positive effect of longer spacing that very short 
spacing---approximately 24 months or less---leads to worse child health and mortality 
outcomes \citep{Whitworth2002,Conde-Agudelo2012}.
Whether longer spacing also improves human capital outcomes is more speculative with
mixed evidence for developed countries and no evidence for developing countries
\citep{Zajonc1976,Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

% Son preference 
Empirically, stronger son preference only marginally increases average fertility, but
differential stopping behavior still affects the sex composition and fertility across 
families and leads to shorter spacing and worse health outcomes for daughters  
when there are no sons in the household
\citep{repetto72,clark00,Whitworth2002,Basu2010,Jayachandran2011,Barcellos2014}.

Stronger son preference increases the use of sex selection and each sex-selective
abortion increases the expected birth spacing by 6--12 months.
After an abortion the uterus needs at least two menstrual cycles---approximately 
two months---to recover, or the subsequent risk of spontaneous abortion increases 
substantially \citep{zhou00b}.
Once conception can be attempted, the waiting time to conception is likely between
one and six months. 
Finally, sex-determination tests are reliable only from three months of gestation. 
Hence, it takes at least six months before the couple is at the same point in the next 
pregnancy as they were in the prior pregnancy when they decided to abort.

The use of sex selection, and therefore the length of birth intervals, increases with
lower desired fertility and with parity, provided the desired number of sons has 
not been reached \citep{Portner2015b,Jayachandran2017}.

% Factors/predictions
Although the explanations above are based on economic conditions, desired child outcomes,
son preference, and sex selection, none of these are available in the data, so I instead 
use female education and area of residence as the two principal explanatory variables in 
the empirical analyses.
Not only are these the closest available proxies, prior research shows that both higher 
female education and urbanization are associated with lower fertility and increased 
sex selection use, and, therefore, likely also changes birth spacing
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Furthermore, both variables are of interest given the tremendous changes in India; 
Figure \ref{fig:education_over_time} shows the substantial increase in female education
and the urban proportion of the population almost doubled from 18\% in 1960 to 35\% in 
2019 \citep{United-Nations2019}.

\begin{figure}[htpb]
\centering
\subfloat[Rural]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_rural}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_urban} 
    \end{minipage}
}
\caption{Distribution of education by cohort for women 20 years or older
using the four rounds of the NFHS}
\label{fig:education_over_time}
\end{figure}

Urban women are likely to have lower fertility and higher use of sex selection than women 
in rural areas.
The lower fertility is the result of higher cost of children and lower child mortality 
risk in urban compared to rural areas.%
\footnote{
However, holding parental education constant, children in Indian slums have worse
health outcomes than in rural areas \citep{Portner2018a}.
}
The higher use of sex selection comes partly from the lower fertility and partly from
higher income and better access to health care facilities in urban areas.

As women receive more education their potential wages increase, which should shorten 
birth spacing and lower fertility, but the low and declining female labor force 
participation in India suggests that families face little incentive to space children 
more closely together for economic reasons.
Even as women's education increased, their labor force participation stagnated or 
decreased and is now one of the lowest outside the Middle East 
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018}.

The decline in female labor force participation appear to be driven by a combination
of a relatively small expansion in the sectors where women work and the income effect 
from rapidly increasing male wages and education dominating the substitution effect from 
higher female wages \citep{Klasen2015,Bhargava2018}.
The income effect can dominate because not only do men have a higher average level of 
education than women, men are also paid more than women for a given level of education, 
and the returns to education increases with education in India \citep{Agrawal2011}.
Hence, increasing male education will result in an larger increase in income than a 
similar increase in female education.

However, in a situation with increasing returns to male education and declining female labor 
force participation, we might still observe an association between higher female education 
and both lower fertility and increased use of sex selection, provided better-educated women 
are better at producing child human capital as suggested by \citet{Behrman1999}.
In this scenario, better-educated women are desired not for their ability to generate
income but for their ability to produce better-educated sons.
The higher return to male education provides an additional incentive to invest in sons'
education---with a better-educated mother one such investment---and also lowers desired 
fertility, makes sex selection more likely, and possibly increases birth spacing.

% Sanskritization
Finally, although the expansion in female education will likely bring in groups who 
initially have different norms, the process of ``Sanskritization'' implies that as 
lower-castes females gain access to education, they adopt higher-caste norms such as stronger son 
preference and a retraction from the formal labor market \citep{Srinivas1956}.
The declining female labor force participation suggests that this process still 
operates \citep{Abraham2013,Chatterjee2018}.

% Summary / hypotheses
In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect longer birth spacing over time, independent of education 
levels.
Furthermore, birth spacing likely increases the most among the better educated 
because their household income increases the most---even with declining female labor 
participation---and because of their use of sex selection.
Finally, "Sanskritization" implies that the changing composition of better-educated
women will not substantially change the use of sex selection.
